---
title: "Assignment - Week 03"
author: "Flurin Schindele"
date: "2026-03-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(rstudioapi)
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

```{r libraries}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(Rsubread) # read alignment
  library(rtracklayer) # read/write genomic file formats
  library(Biostrings) # handling sequences
  library(Rfastp) # read QC/trimming
  library(epiwraps)
})
ah <- AnnotationHub()
```

```{r config}
force_rerun <- TRUE
```

The `force_rerun` flag controls whether compute-intensive steps are re-executed. Set it to `TRUE` to recompute everything from scratch, ignoring any existing output files. With the default value of `FALSE`, each step is skipped if its output file is already present, greatly reducing runtime on repeated knits.

# Processing of ChIP-seq data

## Downloading Drosophila ChIP-seq data

The raw reads are downloaded from the ENCODE portal (accession ENCFF127RRR). The file is only downloaded if it does not already exist locally.

```{r download-raw-reads}
dir.create("data")
dir.create("data/raw")

if (!file.exists("data/raw/ctcf.fastq.gz")) {
  download.file(
    url = "https://www.encodeproject.org/files/ENCFF127RRR/@@download/ENCFF127RRR.fastq.gz",
    destfile = "data/raw/ctcf.fastq.gz",
    mode = "wb"
  )
} else {
  print("File already exists. Skipping download...")
}
```

## Quality control

`Rfastp` performs adapter trimming and quality filtering in a single step. 
It produces a trimmed FASTQ file alongside a JSON report object (`qc`) containing per-read statistics. 
The `qc` object is saved to disk so that the plots below can be rendered even when this step is skipped on re-runs.

```{r qc-trimming}
if (force_rerun || !file.exists("data/raw/trimmed_ctcf_R1.fastq.gz")) {
  qc <- Rfastp::rfastp(
    read1 = "data/raw/ctcf.fastq.gz",
    outputFastq = "data/raw/trimmed_ctcf",
    adapterTrimming = TRUE,
    overrepresentationAnalysis = TRUE,
    thread = 4
  )
  saveRDS(qc, "data/raw/trimmed_ctcf_qc.rds")
} else {
  message("Trimmed reads already exist. Loading cached QC stats.")
  qc <- readRDS("data/raw/trimmed_ctcf_qc.rds")
}
```

The quality curves show per-base quality scores and nucleotide composition before and after trimming.
We have short reads with rather low base quality over all positions but especially towards the end.

```{r qc-curve-plots, out.width="75%", fig.align="center"}
Rfastp::curvePlot(qc, curves = "quality_curves")
```

QC content looks good and duplication levels are also acceptable.
Also, `Rfastp` did not find any adapter sequences present in the reads.

```{r, out.width="75%", fig.align="center"}
Rfastp::curvePlot(qc, curve = "content_curves")
```

The duplication histogram shows how many reads appear at each duplication level.
High duplication levels could indicate PCR over-amplification and potential loss of library complexity but here the distribution looks very good.

```{r qc-duplication-plot, out.width="75%", fig.align="center"}
plot(qc$duplication$histogram,
  type = "b",
  xlab = "Duplication level",
  ylab = "Read count",
  main = "Duplication histogram"
)
```

## Alignment of reads

Before alignment, we need to fetch the *D. melanogaster* BDGP6 genome sequence and build a genome index from the FASTA file.
(The soft-masked version is used, meaning repetitive and low-complexity regions are represented in lowercase letters rather than being replaced with `N` (hard-masking). 
This allows the aligner to down-weight alignments in repetitive regions while still retaining the underlying sequence information.)

```{r download-genome}
if (force_rerun || !file.exists("data/BDGP6_genome/bdgp6.fasta.gz")) {
  dir.create("data/BDGP6_genome", showWarnings = FALSE)

  query_results <- query(
    ah,
    c("drosophila melanogaster", "BDGP6", "dna_sm", "2bit")
  )

  dm_genome <- ah[[tail(rev(query_results$ah_id)[1])]]

  export(import.2bit(dm_genome), "data/BDGP6_genome/bdgp6.fasta.gz", compress = TRUE)
} else {
  message("BDGP6 genome already exists. Skipping download.")
}
```

```{r build-genome-index}
if (force_rerun || !file.exists("data/BDGP6_genome/rsubread.files")) {
  Rsubread::buildindex(
    "data/BDGP6_genome/rsubread",
    reference = "data/BDGP6_genome/bdgp6.fasta.gz"
  )
} else {
  message("Rsubread index already exists. Skipping index building.")
}
```

Trimmed reads are aligned to the BDGP6 genome with `Rsubread::align`.
The output is a coordinate-sorted BAM file. 
Alignment statistics are cached as an RDS file so the Report section can access them even when this step is skipped on re-runs.

```{r align-reads}
dir.create("data/aligned", showWarnings = FALSE)

if (force_rerun || !file.exists("data/aligned/ctcf.bam")) {
  align_stats <- Rsubread::align(
    index = "data/BDGP6_genome/rsubread", type = "dna",
    readfile1 = "data/raw/trimmed_ctcf_R1.fastq.gz",
    output_file = "data/aligned/ctcf.bam",
    sortReadsByCoordinates = TRUE,
    nthreads = 6
  )
  saveRDS(align_stats, "data/aligned/align_stats.rds")
} else {
  message("Aligned BAM already exists. Loading cached alignment stats.")
  align_stats <- readRDS("data/aligned/align_stats.rds")
}
align_stats
```

Enriched genomic regions (peaks) are called from the aligned BAM using MACS3. 
The Drosophila effective genome size (`gsize = "dm"`) is used for the statistical model.

```{r call-peaks}
dir.create("data/peaks", showWarnings = FALSE)

if (!require("MACSr", quietly = TRUE)) BiocManager::install("MACSr")
library(MACSr)

if (force_rerun || !file.exists("data/peaks/ctcf_peaks.narrowPeak")) {
  callpeak(
    tfile = "data/aligned/ctcf.bam",
    gsize = "dm",
    outdir = "data/peaks",
    name = "ctcf"
  )
} else {
  message("Peaks file already exists. Skipping peak calling.")
}

peaks <- rtracklayer::import("data/peaks/ctcf_peaks.narrowPeak", format = "narrowPeak")
peaks <- peaks[order(peaks$score, decreasing = TRUE)]
head(peaks)
```

## Report

1. After filtering by `Rfastp`, we have `r align_stats["Total_reads", "ctcf.bam"]` reads left for alignment. Of those, **`r align_stats["Mapped_reads", "ctcf.bam"]`** were successfully mapped to the reference genome, which corresponds to a mapping rate of **`r round(align_stats["Mapped_reads", "ctcf.bam"] / align_stats["Total_reads", "ctcf.bam"] * 100, 2)`%**.
2. We identified **`r length(peaks)`** peaks in the ChIP-seq data, which are potential binding sites of the CTCF transcription factor in the Drosophila genome.

### Signal around a peak inside a gene

To check what peaks fall inside gene bodies, we first load the latest Ensembl BDGP6 gene annotation from AnnotationHub.
We also need to harmonise Chromosome naming conventions between the annotation and the peaks object, then peaks are subset to those overlapping a gene body.

```{r load-gene-annotations}
query_results <- query(ah, c("Drosophila melanogaster", "EnsDb", "BDGP6"))
dm_ensdb <- ah[[tail(query_results$ah_id, 1)]]

genes <- genes(dm_ensdb)
seqlevelsStyle(genes) <- seqlevelsStyle(peaks) # harmonise chr naming

peaks_in_genes <- suppressWarnings(peaks[overlapsAny(peaks, genes, type = "within")])
```

The signal track below shows read coverage from the CTCF ChIP-seq experiment around a intragenic peak, extended by 1 kb on each side. The gene model is shown below the coverage track.

```{r plot-signal-tracks, out.width="75%", fig.align="center"}
region_of_interest <- peaks_in_genes[1]

overlapping_gene <- suppressWarnings(genes[overlapsAny(genes, region_of_interest, , type = "any")][1]$gene_name)

plotSignalTracks(
  list(ctcf = "data/aligned/ctcf.bam"),
  region = region_of_interest,
  extend = 1000,
  genomeAxis = TRUE,
  ensdb = dm_ensdb,
  transcripts = "full",
  main = paste0("CTCF signal inside ", overlapping_gene, " gene"),
  cex.main = 1.5
)
```

The BAM file is converted to a bigWig coverage track using `bam2bw`. Reads are extended to 176 bp (estimated fragment length by MACS3) before computing coverage.

```{r bam-to-bigwig}
dir.create("data/coverage_tracks", showWarnings = FALSE)

if (force_rerun || !file.exists("data/coverage_tracks/ctcf.bw")) {
  bam2bw(
    bamfile = "data/aligned/ctcf.bam",
    output_bw = "data/coverage_tracks/ctcf.bw",
    extend = 176)
} else {
  message("Coverage track already exists. Skipping bam2bw.")
}
```

A signal matrix is computed centred on the top 200 peaks (ranked by MACS3 score), extended by ±1 kb.
The heatmap visualises ChIP-seq enrichment across all selected peaks, with each row representing one peak.
Here we can see, that the peaks are quite narrow and show a strong enrichment at the center.

```{r signal-heatmap, out.width="75%", fig.align="center"}
sm <- signal2Matrix(
  list(ctcf = "data/coverage_tracks/ctcf.bw"),
  regions = peaks[1:200],
  extend = 1000
)
plotEnrichedHeatmaps(sm)
```
