---
title: "Assignment - Week 05"
author: "Flurin Schindele"
date: "`r Sys.Date()`"
output: html_document
---

<style type="text/css">
  body{
  font-size: 13pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", out.width = "75%")
```

```{r}
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(SummarizedExperiment)
  library(edgeR) # differential gene expression using neg.bin stats
  library(epiwraps)
  library(ggplot2)
  library(sechm) # for plotting (features*samples) heatmaps
  library(dplyr)
})
```

```{r download-data, eval=FALSE}
dir.create("./data", showWarnings = FALSE)
download.file(
  "https://ethz-ins.org/content/mouse_mm38_hippocampus.peakCounts.SE.rds",
  "data/mm38-hippocampus-peaksCounts.rds",
  mode="wb"
)
```

We are working with ATAC-seq data from the hippocampus of mice.
As data we have raw peak counts.
FSS denotes the "forced swim stress" condition.

```{r load-data}
se <- readRDS("data/mm38-hippocampus-peaksCounts.rds")
print(se)
print(colData(se))
```

First, we need to assign labels to the samples.
As we want to compare stressed (FSS) and control samples as well as male and female mice, we need to distinguish both groups.

```{r assign-labels}
colData(se)$condition <- ifelse(grepl("FSS", row.names(colData(se))), "FSS", "Control")
colData(se)$sex <- ifelse(grepl("_male_", row.names(colData(se))), "male", "female")
print(colData(se))
```

We also need to assign row names to the assay.

```{r assign-row-names}
rownames(se) <- as.character(rowRanges(se))
head(rownames(se))
```

# 1. Comparing Sex

For the first analysis, we will focus on the sex differences in chromatin accessibility in the murine hippocampus.
We will be using edgeR for differential analysis.

1. Normalization is done using the TMM method, which should correct for sequencing depth differences between samples and the composition bias.
2. We estimate dispersion in order to correctly model the count data as a negative binomial distribution. The BCV plot can give us some insight into how well this worked.
3. We then perform an exact test to identify differentially accessible regions between male and female mice.

```{r diff-sex}
# Put data into format required by edgeR
dds_sex<- edgeR::DGEList(assay(se), group=se$sex)

# Normalization with TMM
dds_sex <- edgeR::calcNormFactors(dds_sex)
assay(se, "logcpm_sex") <- log1p(edgeR::cpm(dds_sex))

# Dispersion estimation
dds_sex <- edgeR::estimateDisp(dds_sex)

et_sex <- as.data.frame(topTags(exactTest(dds_sex), Inf))
et_sex %>%
  filter(FDR <= 0.05) %>%
  head(10)
```

```{r bcv-plot}
plotBCV(dds_sex)
```

The plot shows us that the biological coefficient of variation (BCV) is relatively high. 
Apparently, the blue line should also decrease with increasing CPM which is not really the case here.
**Question, ask in lecture...**

The table above shows us that the top hits are all located on either the Y or X chromosome.
This should not be surprising for anyone but does not really help us in the analysis.
Maybe this could also influence dispersion estimation.

```{r sex-volcano}
colors <- c("up" = "#C04F15", "down" = "#5682A3", "ns" = "grey") 
sizes <- c("up" = 2, "down" = 2, "ns" = 1) 
alphas <- c("up" = 1, "down" = 1, "ns" = 0.5)

et_sex <- et_sex %>%
  mutate(type = case_when(
    logFC >= 1 & FDR <= 0.05 ~ "up",
    logFC <= -1 & FDR <= 0.05 ~ "down",
    TRUE ~ "ns"
    )
)

et_sex$type <- factor(et_sex$type, levels=c("up", "down", "ns"))

ggplot(data = et_sex,
       aes(x = logFC,
           y = -log10(FDR))) + 
  geom_point(aes(color = type,    
                 size = type,
                 alpha = type)) + 
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed") + 
  geom_vline(xintercept = c(log2(0.5), log2(2)),
             linetype = "dashed") +
  scale_colour_manual(values = colors) +
  scale_size_manual(values = sizes) +
  scale_alpha_manual(values = alphas)
```

We can look further into the count data of the top hits. 
This could show us outliers and should also confirm the sex-chromosome problem.

```{r sex-heatmap}
topRegions_sex <- head(row.names(et_sex), 30)
sechm(se, topRegions_sex, gaps_at="sex", assayName = "logcpm_sex")
```

**Question: What does the do.scale parameter do?**

```{r sex-heatmap-scaled}
sechm(se, topRegions_sex, gaps_at="sex", assayName = "logcpm_sex", do.scale = TRUE)
```

## Filtering out sex chromosomes

By filtering out the sex chromosomes, we can increase the statistical power of our analysis to detect differences on the autosomes.

```{r diff-autosomes}
autosomes <- as.character(c(1:19)) 
se_autosomes <- se[seqnames(se) %in% autosomes, ]

dds_autosomes<- edgeR::DGEList(assay(se_autosomes), group=se$sex)

# Normalization with TMM
dds_autosomes <- edgeR::calcNormFactors(dds_autosomes)
assay(se_autosomes, "logcpm_autosomes") <- log1p(edgeR::cpm(dds_autosomes))

# Dispersion estimation
dds_autosomes <- edgeR::estimateDisp(dds_autosomes)

et_autosomes <- as.data.frame(topTags(exactTest(dds_autosomes), Inf))
head(et_autosomes, 5)
```

```{r autosomes-volcano}
et_autosomes <- et_autosomes %>%
  mutate(type = case_when(
    logFC >= 1 & FDR <= 0.05 ~ "up",
    logFC <= -1 & FDR <= 0.05 ~ "down",
    TRUE ~ "ns"
    )
)

et_autosomes$type <- factor(et_autosomes$type, levels=c("up", "down", "ns"))

ggplot(data = et_autosomes,
       aes(x = logFC,
           y = -log10(FDR))) + 
  geom_point(aes(color = type,    
                 size = type,
                 alpha = type)) + 
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed") + 
  geom_vline(xintercept = c(log2(0.5), log2(2)),
             linetype = "dashed") +
  scale_colour_manual(values = colors) +
  scale_size_manual(values = sizes) +
  scale_alpha_manual(values = alphas)
```

When filtering out the sex chromosomes, we see that there are no significant differences in chromatin accessibility between male and female mice.

# 2. Stressed (FSS) vs. Control

We can repeat the same procedure and compare stressed versus non-stressed mice.

```{r diff-condition}
dds_cond <- edgeR::DGEList(assay(se), group=se$condition)
dds_cond <- edgeR::calcNormFactors(dds_cond)
assay(se, "logcpm_condition") <- log1p(edgeR::cpm(dds_cond))
dds_cond <- edgeR::estimateDisp(dds_cond)

et_cond <- as.data.frame(topTags(exactTest(dds_cond), Inf))
```

```{r condition-volcano}
et_cond <- et_cond %>%
  mutate(type = case_when(
    logFC >= 1 & FDR <= 0.05 ~ "up",
    logFC <= -1 & FDR <= 0.05 ~ "down",
    TRUE ~ "ns"
    )
)

et_cond$type <- factor(et_cond$type, levels=c("up", "down", "ns"))

ggplot(data = et_cond,
       aes(x = logFC,
           y = -log10(FDR))) + 
  geom_point(aes(color = type,    
                 size = type,
                 alpha = type)) + 
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed") + 
  geom_vline(xintercept = c(log2(0.5), log2(2)),
             linetype = "dashed") +
  scale_colour_manual(values = colors) +
  scale_size_manual(values = sizes) +
  scale_alpha_manual(values = alphas)
```

One region is significantly more accessible in stressed mice compared to controls.

```{r}
et_cond %>%
  filter(FDR <= 0.05)
```

```{r condition-heatmap}
topRegions_cond <- head(row.names(et_cond), 20)
sechm(se, topRegions_cond, gaps_at="condition", top_annotation="sex", anno_colors = list("gold", "darkred"), assayName = "logcpm_condition")
```

```{r condition-heatmap-scaled}
# Question: why are anno_colors ignored?
sechm(se, topRegions_cond, gaps_at="condition", top_annotation="sex", anno_colors = list("gold", "darkred"), assayName = "logcpm_condition", do.scale = TRUE)
```

# Extra: Controlling for Sex

We should control for a potential confounding effect of the sex on the condition effect.
For this we can use generalized linear models as described in [edgeR User’s Guide, section 2.12](https://www.bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf).


```{r diff-sex-cond}
# 1. Create design matrix
design <- model.matrix(~ sex + condition, data = colData(se))

# 2. Initialize DGEList and normalize
dge <- DGEList(counts = assay(se))
dge <- calcNormFactors(dge)
assay(se, "logcpm_sex_condition") <- log1p(edgeR::cpm(dge))

# 3. Estimate dispersion using the design matrix
dge <- estimateDisp(dge, design)

# 4. Fit the Quasi-Likelihood GLM (Recommended for better FDR control)
fit <- glmQLFit(dge, design)

# 5. Test for the 'condition' effect which is the last column in the design matrix
qlf <- glmQLFTest(fit, coef = ncol(design))

summary(decideTests(qlf))

res <- as.data.frame(topTags(qlf, n = Inf))
head(res)
```

Sadly, we do not find any significant differences between stressed and control mice when controlling for the sex effect.

```{r mds-plot}
# Create an MDS plot colored by condition and shaped by sex
plotMDS(dge, col = as.numeric(factor(se$condition)), 
        pch = as.numeric(factor(se$sex)))
#legend("topleft", legend = levels(factor(se$condition)), 
#       text.col = 1:nlevels(factor(se$condition)))
```

```{r sessionInfo}
sessionInfo()
```
