---
title: "Assignment - Week 04"
author: "Flurin Schindele"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(rtracklayer)
  library(epiwraps)
  library(GenomicRanges)
})
```


```{r download, eval = FALSE}
dir.create("data", showWarnings = FALSE)

download.file(
  url = "https://www.encodeproject.org/files/ENCFF179FJG/@@download/ENCFF179FJG.bed.gz",
  destfile = "data/p300-mmES-ENCFF179FJG.bed.gz"
)

download.file(
  url = "https://www.encodeproject.org/files/ENCFF360VIS/@@download/ENCFF360VIS.bed.gz",
  destfile = "data/H3K27ac-mmES-ENCFF360VIS.bed.gz"
)

download.file(
  url = "https://www.encodeproject.org/files/ENCFF008XKX/@@download/ENCFF008XKX.bed.gz",
  destfile = "data/H3K27me3-mmES-ENCFF008XKX.bed.gz"
)

download.file(
  url = "https://www.encodeproject.org/files/ENCFF326YGO/@@download/ENCFF326YGO.bed.gz",
  destfile = "data/p300-mmB6NCrlLiver-ENCFF326YGO.bed.gz"
)

download.file(
  url = "https://www.encodeproject.org/files/ENCFF109YUQ/@@download/ENCFF109YUQ.bed.gz",
  destfile = "data/H3K27me3-mmB6NCrlLiver-ENCFF109YUQ.bed.gz"
)

download.file(
  url = "https://www.encodeproject.org/files/ENCFF290MLR/@@download/ENCFF290MLR.bed.gz",
  destfile = "data/H3K27ac-mmB6NCrlLiver-ENCFF290MLR.bed.gz"
)
```

### Data Source and Peak Types

The following BED files were downloaded from the ENCODE portal. Due to differences in available processed data for each experiment, the peak calling strategies vary:

*   **p300 (mESC)**: `ENCFF179FJG` - *Optimal IDR thresholded peaks* (High confidence, consistent across replicates)
*   **H3K27ac (mESC)**: `ENCFF360VIS` - *Replicated peaks* (Peaks present in both biological replicates)
*   **H3K27me3 (mESC)**: `ENCFF008XKX` - *Pseudoreplicated peaks* (Derived from pooling reads and artificially splitting them; used when strict replication isn't available)
*   **p300 (Liver)**: `ENCFF326YGO` - *IDR thresholded peaks* (High confidence)
*   **H3K27me3 (Liver)**: `ENCFF109YUQ` - *Replicated peaks*
*   **H3K27ac (Liver)**: `ENCFF290MLR` - *Replicated peaks*

```{r}
peakfiles <- list.files("data", full.names = TRUE)
names(peakfiles) <- sub("^(.*)-.*\\.bed\\.gz$", "\\1", basename(peakfiles))
peaks <- lapply(peakfiles, FUN=rtracklayer::import, format="narrowPeak")
```

```{r}
ah <- AnnotationHub()

# Query for the latest mouse EnsDb
query_results <- query(ah, c("EnsDb", "Mus musculus"))
ensdb <- ah[[tail(names(query_results), 1)]]
```

## TSS Overlap Analysis

### What do we mean by “TSS”? (gene-level vs transcript-level)

**Genes can have multiple transcripts**, and therefore **multiple possible TSS** (alternative promoters). As explained by Gemini:

- **Gene-level TSS (one per gene):** simple and answers “what fraction of genes have a promoter/TSS overlapping this mark?”. In an EnsDb, the gene range typically spans all isoforms, so this effectively corresponds to the **most 5′ TSS per gene** (per strand).
- **Transcript-level TSS (one per transcript):** closer to what the [HTSeq](https://htseq.readthedocs.io/en/latest/tutorials/tss.html) tutorial shows. It captures alternative promoters, but can **over-weight genes with many isoforms** unless you deduplicate or collapse per gene.

```{r}
# Gene-level TSS

genes_gr <- genes(ensdb)

# ENCODE peak files typically use UCSC seqnames like "chr1".
# EnsDb often uses Ensembl-style seqnames like "1".
# Convert to UCSC so overlaps work.
GenomeInfoDb::seqlevelsStyle(genes_gr) <- "UCSC"

# 1bp at the gene-level TSS
# (one per gene; roughly the most 5′ start across isoforms, per strand)

# promoters() is strand-aware: on '+' the TSS is at the start of the range,
# on '-' it is at the end. promoters(x, upstream=0, downstream=1) returns a 1bp
# GRanges exactly at the transcription start site definition.
tss_gene <- promoters(genes_gr, upstream = 0, downstream = 1)

length(tss_gene)
```

```{r}
# Transcript-level TSS

# One per transcript (can include multiple distinct TSS per gene)
tx_gr <- transcripts(ensdb)
GenomeInfoDb::seqlevelsStyle(tx_gr) <- "UCSC"

tss_tx <- promoters(tx_gr, upstream = 0, downstream = 1)

# Avoid counting identical TSS positions multiple times
# (still: genes with multiple distinct promoters will appear multiple times)
tss_tx_unique <- unique(tss_tx)

c(n_tx = length(tss_tx), n_tx_unique = length(tss_tx_unique))
```

Now we can calculate the proportion of TSS overlapping with our peaks.

```{r, warning=FALSE}
# Helper function to calculate proportion of TSS overlapping peaks
calc_overlap <- function(peak_set, tss_set) {
  hits <- findOverlaps(tss_set, peak_set) # this returns a Hits object
  length(hits) / length(tss_set) # we assume that all query hits are unique
}

# Proportion of TSS overlapping BOTH peak sets (logical AND)
calc_overlap_both <- function(tss_set, peak_a, peak_b) {
  in_a <- overlapsAny(tss_set, peak_a) # return logical vector of length len(tss_set)
  in_b <- overlapsAny(tss_set, peak_b)
  # as in_a and in_b are of length len(tss_set), 
  # we can simply calculate the mean over the logical vector
  mean(in_a & in_b) 
}

# Apply to all our peak sets (gene-level TSS)
overlap_gene_tss <- sapply(peaks, calc_overlap, tss_set = tss_gene)

# Apply to all our peak sets (transcript-level TSS; unique positions)
overlap_tx_tss <- sapply(peaks, calc_overlap, tss_set = tss_tx_unique)

res1 <- rbind(
  gene_level_TSS = overlap_gene_tss,
  transcript_level_TSS_unique = overlap_tx_tss
)

# Add combined overlaps for H3K27ac AND H3K27me3 in each tissue
res1 <- cbind(
  res1,
  `H3K27mixed-mmB6NCrlLiver` = c(
    gene_level_TSS = calc_overlap_both(
      tss_gene,
      peaks$`H3K27ac-mmB6NCrlLiver`,
      peaks$`H3K27me3-mmB6NCrlLiver`
    ),
    transcript_level_TSS_unique = calc_overlap_both(
      tss_tx_unique,
      peaks$`H3K27ac-mmB6NCrlLiver`,
      peaks$`H3K27me3-mmB6NCrlLiver`
    )
  ),
  `H3K27mixed-mmES` = c(
    gene_level_TSS = calc_overlap_both(
      tss_gene,
      peaks$`H3K27ac-mmES`,
      peaks$`H3K27me3-mmES`
    ),
    transcript_level_TSS_unique = calc_overlap_both(
      tss_tx_unique,
      peaks$`H3K27ac-mmES`,
      peaks$`H3K27me3-mmES`
    )
  )
)
```

### 1. Proportion of TSS that overlap epigenetic marks

The tables below show the proportion of the peaks overlapping either the TSS we have defined on the gene or transcript level.

In the embryonic stem cells, roughly $2 \percent$ of TSS are marked with H3K27ac. 
When looking at transcript level TSS, this proportion is slightly higher. 
This could be due to the proximity of TSS for transcripts from the same gene and the fact that these histone modifications can be relatively "dispersed" (not extremely localized) over a genomic region and therefore overlapping multiple TSS at the same time.
H3K27-trimethylation is much rarer, and overlap of both marks wasn't observed.
This could hint at the uniformity of the cellular phenotype in the experiment, comparted to a differentiated organ (see below).

```{r}
res1[, c("H3K27ac-mmES", "H3K27me3-mmES", "H3K27mixed-mmES")]
```

As in stem cells, more TSS are marked with H3K27-acetylation rather than trimethylation in the liver.
This indicates that roughly $2.5 \times$ more TSS are active than repressed.
The proportion of "poised" or bivalent TSS is quite low but higher than in stem cells.
As the liver contains multiple cell types, they show different gene regulation marks due to their diverse phenotypes.

```{r}
res1[, c("H3K27ac-mmB6NCrlLiver", "H3K27me3-mmB6NCrlLiver", "H3K27mixed-mmB6NCrlLiver")]
```
