---
title: "Assignment - Week 08"
author: "Flurin Schindele"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: true
    highlight: default
    df_print: paged
---

<style type="text/css">

body{
  font-size: 13pt;
}

/* Main container - set position for TOC alignment */
.main-container {
    max-width: 940px;
    margin-left: auto;
    margin-right: auto;
}

/* TOC box styling - elegant and light */
.tocify {
  width: 100%;
  max-height: calc(100vh - 100px);
  overflow-y: auto;
  background: #fff;
  padding: 10px 5px;
  border: none;
  border-left: 1px solid #ddd;
  border-radius: 0;
  box-shadow: none;
  pointer-events: auto;
}

.tocify ul, .tocify li {
  list-style: none;
  margin: 0;
  padding: 0;
}

.tocify-header {
  text-indent: 0;
}

.tocify-subheader {
  font-size: 90%;
}

.tocify a {
  color: inherit;
  text-decoration: none;
  display: block;
  cursor: pointer;
}

.tocify .list-group-item:hover {
  background-color: transparent;
  color: #555;
  border-left: 2px solid #aaa;
}

.tocify .list-group-item.active {
  background-color: transparent;
  color: #333;
  border-left: 2px solid #555;
  font-weight: 500;
}

.tocify .list-group-item {
  background-color: transparent;
  border: none;
  padding: 2px 8px 2px 10px;
  margin: 0;
  display: inline-block;
  color: #888;
  border-left: 2px solid transparent;
  margin-left: -1px;
  cursor: pointer;
  line-height: 1.3;
  /* Text wrapping for long headers */
  white-space: normal;
  word-wrap: break-word;
  overflow-wrap: break-word;
}

/* Subheader indentation */
.tocify-subheader .list-group-item {
  padding-left: 5px;
}

p.caption {
  margin: 0 10px 0 10px;
  font-style: italic;
  font-size: 90%;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", out.width = "75%")
```

```{r packages, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(epiwraps)
  library(ggplot2)
  library(AnnotationHub)
  library(ensembldb)
  library(rtracklayer)
  library(GenomicRanges)
  library(GenomeInfoDb)
  # library(edgeR)
  library(BiocParallel)
  library(bsseq)   # Class and tools to analyze bisulfite sequencing data at CpG-level
  library(DMRcate) # Identification of differentially-methylated regions
})
```

# Plotting DNAme of genes in top 5 DMR

We first load the differentially methylation region data, as well as BSseq object containing the methlyation signal. We also apply smoothing like in the tutorial. 

```{r}
dmr <- readRDS("data/dmr.rds")
head(dmr)
```

```{r}
if (!file.exists("data/ESC_astrocytes_BSseq_smoothed.rds")) {
  bs <- readRDS("data/ESC_astrocytes_BSseq.rds")
  bs <- bs[rowSums(assay(bs,"Cov")>0)==ncol(bs),]
  bs <- BSmooth(bs, BPPARAM=MulticoreParam(4, progress=FALSE))
  saveRDS(bs, "data/ESC_astrocytes_BSseq_smoothed.rds")
} else {
  bs <- readRDS("data/ESC_astrocytes_BSseq_smoothed.rds")
}
bs
```

We subset to chromosome 1 and select the top 5 DMRs by HMFDR, then find genes that overlap these regions and plot their methylation levels across samples.

```{r}
# get regions from chromosome 1
roi <- dmr[seqnames(dmr) == "chr1"]
roi <- roi[order(roi$HMFDR)][1:5]
head(roi)

# get genes that overlap regions of interest
genes <- readRDS("data/hg19_chr1_genes.rds")
genes_oi <- subsetByOverlaps(genes, roi, ignore.strand=TRUE, type="any")
genes_oi
```

```{r}
metROI <- bsseq::getMeth(bs, regions=genes_oi, what="perRegion")
rownames(metROI) <- genes_oi$symbol

pheatmap(metROI, 
         name = "Methylation", 
         main = "Methylation Levels of Genes in Top 5 DMRs (Chr1)",
         show_rownames = TRUE,
         color = colorRampPalette(c("blue", "white", "red"))(50))
```

Interestingly, this approach only shows on gene. The top 5 DMRs also show 2 other genes but they were not found in the hg19 genes.
We will try again with a larger set of DMRs.

```{r}
# get regions from chromosome 1
roi20 <- dmr[seqnames(dmr) == "chr1"]
roi20 <-roi20[order(roi20$HMFDR)][1:20]
tail(roi20)

# get genes that overlap regions of interest
genes_oi20 <- subsetByOverlaps(genes, roi20, ignore.strand=TRUE, type="any")
genes_oi20$symbol
```

```{r}
metROI20 <- bsseq::getMeth(bs, regions=genes_oi20, what="perRegion")
rownames(metROI20) <- genes_oi20$symbol

pheatmap(metROI20, 
         name = "Methylation", 
         main = "Methylation Levels of Genes in Top 20 DMRs (Chr1)",
         show_rownames = TRUE,
         color = colorRampPalette(c("blue", "white", "red"))(50))
```

The first few regions show very distinct methylation profiles between conditions witih ESCs showing higher methylation levels than astrocytes.

# Agreement between DNAme and ATAC

We first load the data and then rebuild the object as there seems to be some bug present.

```{r}
atac_path <- "data/ATAC_AstroVsESC_subset.rds"
if (!file.exists(atac_path)) {
  download.file(
    "https://www.ethz-ins.org/content/ATAC_AstroVsESC_subset.rds",
    destfile = atac_path,
    mode = "wb"
  )
}
atac_da_raw <- readRDS(atac_path)

# Rebuild as a plain GRanges to avoid legacy Seqinfo class issues across Bioc versions.
atac_df <- as.data.frame(atac_da_raw)
atac_da <- GRanges(
  seqnames = atac_df$seqnames,
  ranges = IRanges(atac_df$start, atac_df$end),
  strand = atac_df$strand
)
mcols(atac_da) <- S4Vectors::DataFrame(
  atac_df[, setdiff(colnames(atac_df), c("seqnames", "start", "end", "width", "strand")), drop=FALSE]
)
atac_da
```

We then query the gene annotations from Ensembl, subset to protein-coding genes, and define promoter and gene body regions. We will then compare the DMRs and ATAC peaks that overlap these regions, and compute agreement metrics.

```{r}
ah <- AnnotationHub()
ensdb <- ah[["AH119325"]]
genes <- genes(ensdb)[genes(ensdb)$gene_biotype == "protein_coding"]
genes <- keepStandardChromosomes(genes, pruning.mode="coarse")
seqlevelsStyle(genes) <- "UCSC"

promoters <- promoters(genes, upstream=2000, downstream=500)
gene_bodies <- genes
```

All ATAC peaks overlap with DMRs and roughly 97% of DMRs overlap with ATAC peaks, which is a very high degree of overlap.

```{r}
prop_overlap <- function(query_peaks, subject_peaks) {
  mean(overlapsAny(query_peaks, subject_peaks))
}
prop_overlap(atac_da, dmr)
prop_overlap(dmr, atac_da)
```

```{r}
summarize_region_agreement <- function(region_gr, region_label, dmr_gr, atac_gr) {
  # first get signals that overlap the regions of interest (promoters or gene bodies)
  dmr_sub <- subsetByOverlaps(dmr_gr, region_gr, ignore.strand=TRUE)
  atac_sub <- subsetByOverlaps(atac_gr, region_gr, ignore.strand=TRUE)
  # we also get the overlaps between the DMRs and ATAC peaks to compare their signal directly
  ov <- findOverlaps(dmr_sub, atac_sub, ignore.strand=TRUE)
  
  if (length(ov) == 0) {
    return(list(
      comp = data.frame(),
      metrics = data.frame(
        region_type = region_label,
        overlap_coefficient_pct = NA_real_,
        pearson_cor = NA_real_,
        spearman_cor = NA_real_
      )
    ))
  }
  
  ov_df <- data.frame(
    dmr_ix = queryHits(ov),
    # get mean difference in methylation and HMFDR for the DMRs that overlap ATAC peaks
    dmr_meandiff = dmr_sub$meandiff[queryHits(ov)],
    dmr_HMFDR = dmr_sub$HMFDR[queryHits(ov)],
    # get logFC and FDR for the ATAC peaks that overlap DMRs
    atac_logFC = atac_sub$logFC[subjectHits(ov)],
    atac_FDR = atac_sub$FDR[subjectHits(ov)]
  )
  
  dmr_comp <- data.frame(
    dmr_ix = as.integer(names(tapply(ov_df$atac_logFC, ov_df$dmr_ix, mean, na.rm=TRUE))),
    # extract mean atac logFC for each DMR by averaging across all overlapping ATAC peaks
    atac_logFC_mean = as.numeric(tapply(ov_df$atac_logFC, ov_df$dmr_ix, mean, na.rm=TRUE)),
    atac_FDR_min = as.numeric(tapply(ov_df$atac_FDR, ov_df$dmr_ix, min, na.rm=TRUE)),
    n_atac_peaks = as.integer(table(ov_df$dmr_ix))
  )
  
  dmr_comp$dmr_meandiff <- dmr_sub$meandiff[dmr_comp$dmr_ix]
  dmr_comp$dmr_HMFDR <- dmr_sub$HMFDR[dmr_comp$dmr_ix]
  dmr_comp$region_type <- region_label
  # If both methylation and accessibility changes are in the same direction (both positive or both negative), we call it "agree", otherwise "disagree"
  dmr_comp$signal_change <- ifelse(sign(dmr_comp$dmr_meandiff) == sign(dmr_comp$atac_logFC_mean),
                               "agree", "disagree")
  
  # Calculate both pearson and spearman correlation between methylation changes and mean ATAC logFC across overlapping regions
  corr_pearson <- cor(dmr_comp$dmr_meandiff, dmr_comp$atac_logFC_mean, method="pearson")
  corr_spearman <- cor(dmr_comp$dmr_meandiff, dmr_comp$atac_logFC_mean, method="spearman")
  
  list(
    comp = dmr_comp,
    metrics = data.frame(
      region_type = region_label,
      pearson_cor = corr_pearson,
      spearman_cor = corr_spearman
    )
  )
}

promoter_res <- summarize_region_agreement(promoters, "Promoters", dmr, atac_da)
gene_body_res <- summarize_region_agreement(gene_bodies, "Gene bodies", dmr, atac_da)
region_comp <- rbind(promoter_res$comp, gene_body_res$comp)

region_comp$signal_change <- factor(region_comp$signal_change, levels=c("agree", "disagree"))
region_comp$region_type <- factor(region_comp$region_type, levels=c("Promoters", "Gene bodies"))
region_comp
```

```{r}
ggplot(region_comp, aes(x=dmr_meandiff, y=atac_logFC_mean, color=signal_change, size=n_atac_peaks)) +
  geom_hline(yintercept=0, linetype="dashed", color="grey70") +
  geom_vline(xintercept=0, linetype="dashed", color="grey70") +
  geom_point(alpha=0.8) +
  scale_color_manual(values=c(agree="#1b9e77", disagree="#d95f02")) +
  facet_wrap(~region_type) +
  labs(
    x="DMR methylation change (meandiff; astrocyte vs ESC)",
    y="Mean ATAC logFC of overlapping peaks (astrocyte vs ESC)",
    color="Direction",
    size="# overlapping\nATAC peaks",
    title="Comparison of differential methylation and accessibility"
  ) +
  theme_classic()
```

The plots shos us that there seems to be some disagreement between methlation and accessibility both at the promoters where we would expect it and the gene bodies as well.

In gene bodies, the opposite effect of methylation and accessibility is more pronounced, with most points showing decreasing methylation co-occurring with increased accessibility. At promoters, there is also a tendency for opposite-direction changes, but there are more points showing agreement between methylation and accessibility changes compared to gene bodies.

```{r}
agree_summary <- as.data.frame(prop.table(table(region_comp$region_type, region_comp$signal_change), 1))
colnames(agree_summary) <- c("region_type", "direction", "fraction")
agree_summary
```

```{r}
region_metrics <- rbind(promoter_res$metrics, gene_body_res$metrics)
region_metrics
```

Both correlation metrics show a negative correlation of methlyation and accessibility, which hints at (limited) exclusivity of signals.
