---
title: "ATAC-seq exploration practical"
author: "Pierre-Luc"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(epiwraps)
  library(AnnotationHub)
  library(MotifDb)
  #library(memes)
  library(universalmotif)
  library(ensembldb)
  library(ggplot2)
})

ah <- AnnotationHub(localHub=TRUE)
ensdb <- ah[["AH89211"]] # mouse ensembldb object
```


## Download the data

You can either start from the bam file:

```{r, eval=FALSE}
download.file("https://ethz-ins.org/content/w7/atac.chr19.bam", "atac.chr19.bam", mode="wb")
download.file("https://ethz-ins.org/content/w7/atac.chr19.bam.bai", "atac.chr19.bam.bai", mode="wb")
```

Alternatively, if you're on a slow computer, you can also download the tracks that we're generating below, and then skip that step:
```{r, eval=TRUE}
dir.create("data", showWarnings = FALSE)
download.file("https://ethz-ins.org/content/w7/w7_bw.zip", "data/w7_bw.zip")
unzip("data/w7_bw.zip", exdir="data")
```


## Preparing tracks

```{r}
bam <- "data/atac.chr19.bam"

# create a track using all fragments, computing the coverage at each position (and force Ensembl naming)
epiwraps::bam2bw(bam, output_bw = "data/full_cov.bw", paired=TRUE, binWidth=10L, forceSeqlevelsStyle = "Ensembl")

# create a track using all fragments, but computing the number of cuts/insertion sites at each position
bam2bw(bam, output_bw = "data/full_cuts.bw", paired=TRUE, binWidth=1L, type="ends", 
                 trim=4L, extend=3L, forceSeqlevelsStyle = "Ensembl")

# create a track using only nucleosome-free fragments, computing coverage
bam2bw(bam, output_bw = "data/NF_cov.bw", paired=TRUE, binWidth=5L, minFragLength=30, 
       maxFragLength=120, forceSeqlevelsStyle = "Ensembl")

# create a track using only nucleosome-free fragments, the number of cuts/insertion sites at each position
bam2bw(bam, output_bw = "data/NF_cuts.bw", paired=TRUE, binWidth=1L, type="ends", minFragLength=30, 
       maxFragLength=120, trim=4L, extend=3L, forceSeqlevelsStyle = "Ensembl")

# create a track using only mono-nucleosome fragments, computing coverage
bam2bw(bam, output_bw = "data/mono_cov.bw", paired=TRUE, binWidth=10L, minFragLength=140,
       maxFragLength=220, forceSeqlevelsStyle = "Ensembl")

# create a track using only the (10bp) centers of mono-nucleosome fragments
bam2bw(bam, output_bw = "data/mono_centers.bw", paired=TRUE, binWidth=5L, minFragLength=140, trim=4L,
       maxFragLength=220, type="center", extend=10L, forceSeqlevelsStyle = "Ensembl")
```


## Obtaining the sites with a CTCF motif

```{r}
# get CTCF motif
motif <- MotifDb::query(MotifDb, c("CTCF","Mus"))[[1]]
motif2 <- convert_motifs(motif, class="TFBSTools-PFMatrix")
genome <- ah[["AH68356"]]
# get the sequence for chr19:
chr19 <- import(genome)["19"]

# find motif matches across chr19
moi <- motifmatchr::matchMotifs(motif2, chr19, out="positions", p.cutoff=1e-5)[[1]]
# convert to GRanges
moi <- as(setNames(moi,names(chr19)), "GRanges")
```

## Plotting signal around the motif occurences

```{r}
# we prepare the list of tracks
tracks <- list.files(path=".", recursive=TRUE, pattern="bw$")
names(tracks) <- gsub("\\.bw","",basename(tracks))

# extract signals around the motif occurences
# we zoom in to 300bp around the motif centers, in windows of 5bp
sm <- signal2Matrix(tracks, moi, w=5, extend=300)

# plot the signals:
plotEnrichedHeatmaps(sm, trim=0.95, multiScale = TRUE, axis_name = "0", use_raster=TRUE, raster_device="png")
```


Mono centers:
no enrichment of mono-center reads around CTCF motifs, but sharp peaks to left and right -> well positioned nucleosomes
Nucleosome can move on DNA, slide along. When CTCF binds, it can move/displace the nucleosome 
well-positioned nucleosome is in the same position over different cells with same state

NF cuts: 
TF footprint shown by signal free region
degree depends on binding strength of TF to DNA, stronger/longer binding (long residence time)-> stronger footprint which is the case for CTCF, binding over multiple seconds into minutes which is really long
often TF binds for milliseconds, after this time Tn5 can insert and create a cut -> weak footprint

## Summarizing the signals

To better visualize the relationship between signals, we can overlay them aggregated across regions:

```{r}
d <- meltSignals(sm[,c("NF_cuts", "mono_centers")])
ggplot(d, aes(position, mean, colour=sample)) + geom_line(size=1.2)
```

## Signal around TSS

We could also look at the pattern around the TSS

```{r, fig.width=10, fig.height=6}
# fetch the protein-coding TSS from chromosome 19
txs <- transcripts(ensdb, filter=TxBiotypeFilter("protein_coding"))
txs <- txs[seqnames(txs)=="19"]
tss <- reduce(promoters(txs, upstream=0, downstream=1))
tss.sm <- epiwraps::signal2Matrix(tracks, granges(tss), w=10, extend=1000)
epiwraps::plotEnrichedHeatmaps(tss.sm, trim=0.95, colors=c("white","darkred"), multiScale=TRUE, axis_name="TSS")
```

