---
title: "Assignment - Week 09"
author: "Flurin Schindele"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: true
    highlight: default
    df_print: paged
---

<style type="text/css">

body{
  font-size: 13pt;
}

/* Main container - set position for TOC alignment */
.main-container {
    max-width: 940px;
    margin-left: auto;
    margin-right: auto;
}

/* TOC box styling - elegant and light */
.tocify {
  width: 100%;
  max-height: calc(100vh - 100px);
  overflow-y: auto;
  background: #fff;
  padding: 10px 5px;
  border: none;
  border-left: 1px solid #ddd;
  border-radius: 0;
  box-shadow: none;
  pointer-events: auto;
}

.tocify ul, .tocify li {
  list-style: none;
  margin: 0;
  padding: 0;
}

.tocify-header {
  text-indent: 0;
}

.tocify-subheader {
  font-size: 90%;
}

.tocify a {
  color: inherit;
  text-decoration: none;
  display: block;
  cursor: pointer;
}

.tocify .list-group-item:hover {
  background-color: transparent;
  color: #555;
  border-left: 2px solid #aaa;
}

.tocify .list-group-item.active {
  background-color: transparent;
  color: #333;
  border-left: 2px solid #555;
  font-weight: 500;
}

.tocify .list-group-item {
  background-color: transparent;
  border: none;
  padding: 2px 8px 2px 10px;
  margin: 0;
  display: inline-block;
  color: #888;
  border-left: 2px solid transparent;
  margin-left: -1px;
  cursor: pointer;
  line-height: 1.3;
  /* Text wrapping for long headers */
  white-space: normal;
  word-wrap: break-word;
  overflow-wrap: break-word;
}

/* Subheader indentation */
.tocify-subheader .list-group-item {
  padding-left: 5px;
}

p.caption {
  margin: 0 10px 0 10px;
  font-style: italic;
  font-size: 90%;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", out.width = "75%")
```

```{r libraries}
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(epiwraps)
  library(ggplot2)
  library(rGREAT) # Gene Ontology enrichment among genomic regions
})
```

```{r}
dir.create("data", showWarnings = FALSE)
options(timeout = 6000)
download.file(
  url = "https://ethz-ins.org/content/clustering_assignment.zip",
  destfile = "data/clustering.zip")
unzip("data/clustering.zip", exdir = "data")
```

```{r}
tracks <- list.files(path="data", pattern = ".bw$", full.names = TRUE)
peaks <- lapply(list.files(path="data", pattern = ".bed$", full.names = TRUE), rtracklayer::import.bed)
peaks <- lapply(peaks, FUN=function(x) x[x$score>800])
# concat the individual regions to one GRanges object:
regions_concat <- reduce(unlist(GRangesList(peaks)))
```

```{r}
ese <- signal2Matrix(tracks, regions_concat, extend=2000)
plotEnrichedHeatmaps(ese, cluster_rows = FALSE)
```

```{r}
set.seed(42)
cl2 <- clusterSignalMatrices(ese, k=2:10)
ggplot(cl2$varExplained, aes(k, varExplained)) + geom_line()
```

```{r}
rowData(ese)$cluster <- cl2$clusters[[4]]
head(rowData(ese))
```

```{r}
# color palette for 5 clusters:
mycolors <-wesanderson::wes_palette("Royal2", 5)
names(mycolors) <- c("1", "2", "3", "4", "5")
plotEnrichedHeatmaps(ese, row_split="cluster", mean_color=mycolors, colors=c("white","darkred"))
```

```{r}
d <- meltSignals(ese, splitBy=cl2$clusters[[4]])
ggplot(d, aes(position, mean, colour=sample)) + geom_line(linewidth = 1.2) + facet_wrap(~split)
```

Cluster 2 and 3 look very similar, with slightly different magnitudes of the signal.
This should be taken into consideration when doing other analyses and we could try clustering based on the scaled row data.
The signals from Creb1 and Creb3L1 are pretty highly correlated, whereas Creb3 has a different pattern.

```{r}
split_regions <- split(rowRanges(ese), rowData(ese)$cluster)
lengths(split_regions)

res <- great(split_regions[["1"]], gene_sets="GO:BP", tss_source="hg38", 
             background=regions_concat, cores=8)
bp <- getEnrichmentTables(res)
ggplot(head(bp,15), aes(fold_enrichment, reorder(description, p_adjust), 
                        size=observed_region_hits, color=-log10(p_adjust))) + 
  geom_point() + scale_color_viridis_c() +
  scale_y_discrete(labels = function(x) stringr::str_wrap(x, width = 45)) +
  ggtitle("Creb3 High - Cluster 1 vs all")
```

Cluster 1 with high Creb3 signal is enriched for terms related to responses to general stimuli as well as stress.
Creb3 seems to be involved in communication and signaling as well.
There is also an enriched proliferation/multicellular organisation pattern visible.
This is somewhat in line with the [description](https://www.genecards.org/cgi-bin/carddisp.pl?gene=CREB3&keywords=creb3) of the protein, as it is ER bound and then cleaved in response to cellular stress
which then allows it to translocate to the nucleus.


```{r}
res_2_3 <- great(split_regions[["2"]], gene_sets="GO:BP", tss_source="hg38", 
             background=split_regions[["3"]], cores=4, verbose=FALSE)
bp_2_3 <- getEnrichmentTables(res_2_3)
nrow(bp_2_3)
```

When we check for biological processes that might be enriched between cluster 2 and 3, we find no significant terms.
This is not surprising, as the two clusters look very similar in terms of their signal patterns and their difference might arise
from technical artifacts.
This could also be due to the small size of the sets of regions we are comparing in this case.
We can try to compare the two clusters to the entire background of all regions.

```{r}
res_23 <- great(c(split_regions[["2"]], split_regions[["3"]], ignore.mcols=FALSE), gene_sets="GO:BP", tss_source="hg38", 
             background=regions_concat, cores=4, verbose=FALSE)
bp_23 <- getEnrichmentTables(res_23)

ggplot(head(bp_23), aes(fold_enrichment, reorder(description, p_adjust), 
                        size=observed_region_hits, color=-log10(p_adjust))) + 
  geom_point() + scale_color_viridis_c() +
  scale_y_discrete(labels = function(x) stringr::str_wrap(x, width = 45)) +
  ggtitle("Cluster 2 and 3 vs all - Creb1 + Creb3L1 high")
```

When we compare cluster 2 and 3 to the entire background, we find some significant terms
but they are not very specific.
Multiple small gene sets related to germ cells seem to be enriched and there are also some terms related to RNA processing and translation.
